A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-25 arxiv/transfer.tex
wip

Based on the above intuition, in the following we develop generalization bounds that relate the bias and variance of multi-task learning to the sample sizes and the covariance matrices of both tasks.

@@

\begin{theorem}[Two-task case]\label{thm_main_informal}
	For the setting of two tasks, let $\delta > 0$ be a fixed error margin, $\rho_2 > 1$ and $\rho_1 \gtrsim \delta^{-2}\cdot \lambda_{\min}(M)^{-4} \norm{\Sigma_1} \max(\norm{\beta_1}^2, \norm{\beta_2}^2)$.
	Let $B, W_1, W_2$ be any local minimum of equation \eqref{eq_mtl}.
 	There exist two deterministic functions $\Delta_{\bias}$ and $\Delta_{\vari}$ that only depend on scaled model distance $\beta_1 - \frac{W_1}{W_2} \beta_2$, sample sizes $\rho_1, \rho_2$, and covariate shift matrix $M$ such that
	\begin{enumerate}
		\item[a)] \textbf{Positive transfer:} If $\Delta_{\bias} < \Delta_{\vari} -\delta$, then w.h.p. over the randomness of $X_1, X_2$, we have
			\[ \te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL}).  \]
		\item[b)] \textbf{Negative transfer:} If $\Delta_{\bias} > \Delta_{\vari} + \delta$, then w.h.p. over the randomness of $X_1, X_2$, we have
			\[ \te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL}). \]
	\end{enumerate}
\end{theorem}

@@

\paragraph{Proof overview.}
We shall refer to random matrices of the form $X^\top X$ as sample covariance matrices following the standard notations in high-dimensional statistics. The following two lemmas  are the main random matrix theoretical results of this paper, which will be used to estimate the two terms on the right-hand side of \eqref{eq_te_mtl_2task}.
%which deals with the inverse of the sum of two random matrices, which
%any is can be viewed as a special case of Theorem \ref{thm_model_shift}.

\begin{lemma}[Variance bound]\label{lem_cov_shift}
	%Let $X_i\in\real^{n_i\times p}$ be a random matrix that contains i.i.d. row vectors with mean $0$ and variance $\Sigma_i\in\real^{p\times p}$, for $i = 1, 2$.
	Suppose $X_1=Z_1\Sigma_1^{1/2}\in \R^{n_1\times p}$ and $X_2=Z_2\Sigma_2^{1/2}\in \R^{n_2\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho_1:=n_1/p>1$ and $\rho_2:=n_2/p>1$ being fixed constants.
	Denote by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in descending order. Let $A$ be any $p\times p$ matrix that is independent of $X_1$ and $X_2$. We have that for any constant $\e>0$,
	%When $n_1 = c_1 p$ and $n_2 = c_2 p$, we have that with high probability over the randomness of $X_1$ and $X_2$, the following equation holds
	\begin{align}\label{lem_cov_shift_eq}
		\bigtr{(\hat v^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}A} = \frac{1}{\rho_1+\rho_2}\frac1p\bigtr{ (\hat v^2 a_1 \Sigma_1 + a_2\Sigma_2)^{-1} A} +\bigo{\|A\| p^{-1/2+\epsilon}}
	\end{align}
with high probability, where $(a_1, a_2)$ is the solution to the following deterministic equations:
	\begin{align}
		a_1 + a_2 = 1- \frac{1}{\rho_1 + \rho_2},\quad a_1 + \frac1{\rho_1 + \rho_2}\cdot \frac{1}{p}\sum_{i=1}^p \frac{\hat v^2 \lambda_i^2 a_1}{\hat v^2 \lambda_i^2 a_1 + a_2} = \frac{\rho_1}{\rho_1 + \rho_2}. \label{eq_a12extra}
	\end{align}
\end{lemma}

Lemma \ref{lem_cov_shift} allows us to get a tight bound on equation \eqref{eq_te_var}, that only depends on \textit{sample size}, \textit{covariate shift} and the scalar $\hat{v}$.
As a remark, the concentration error $\bigo{\|A\| p^{-1/2+\epsilon}}$ of our result is nearly optimal.
For the bias term of equation \eqref{eq_te_model_shift}, a similar result that scales with task model distance in addition to sample size and covariate shift holds (cf. Lemma \ref{lem_cov_derivative} in Appendix \ref{app_proof_main_thm}).

\begin{lemma}[Bias bound]\label{lem_cov_derivative}
In the setting of Lemma \ref{lem_cov_shift}, let $\beta \in \R^p$ be any vector that is independent of $X_1$ and $X_2$. We have that for any constant $\e>0$,
\begin{equation}\label{lem_cov_derv_eq}
\begin{split}
&(n_1+n_2)^2\bignorm{\Sigma_2^{1/2} (\hat v^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\beta}^2 \\
&= \beta^{\top} \Sigma_2^{-1/2}  \frac{(1 + a_3)\id + \hat v^2a_4 {M}^{\top}{M}}{( \hat v^2a_1 {M}^{\top}{M}+a_2)^2} \Sigma_2^{-1/2} \beta +\OO(p^{-1/2+\e}\|\beta\|^2),
\end{split}
\end{equation}
with high probability, where $a_{3}$ and $a_4$ satisfy the following system of linear equations:
\begin{gather}\label{eq_a34extra}
		\left(\rho_2  a_2^{-2}-  b_0\right)\cdot  a_3 - b_1 \cdot  a_4
		= b_0, \quad \left(\rho_1 a_1^{-2} -  b_2  \right)\cdot  a_4 -  b_1 \cdot  a_3 = b_1 .
%		\left(\frac{n_1}{\hat a_1^2} -  \sum_{i=1}^p \frac{\hat \lambda_i^4   }{  (\hat a_2 + \hat \lambda_i^2\hat a_1)^2  }\right)\hat a_4 -\left(\sum_{i=1}^p \frac{\hat \lambda_i^2  }{  (\hat a_2 + \hat \lambda_i^2\hat a_1)^2  }\right)\hat a_3
%		= \sum_{i=1}^p \frac{\hat \lambda_i^2 }{  (\hat a_2 + \hat \lambda_i^2\hat a_1)^2  }. \label{eq_a4}
	\end{gather}
Here $b_0$, $b_1$ and $b_2$ are defined as
$$ b_k:= \frac1{p}\sum_{i=1}^p \frac{ \hat v^{2k}\lambda_i^{2k}}{ ( a_2 + \hat v^2\lambda_i^2 a_1)^2  },\quad k=0,1,2.$$
\end{lemma}

\textbf{Proof overview.}
%We first describe the proof of Theorem \ref{lem_cov_shift_informal}.
The proofs of Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative} are based on the Stieltjes transform method (or the resolvent method) in random matrix theory \cite{bai2009spectral,tao2012topics,erdos2017dynamical}. Roughly speaking, we study the resolvent $R(z):=[\Sigma_2^{-1/2}( X_1^{\top}X_1 + X_2^{\top}X_2)\Sigma_2^{-1/2}-z]^{-1}$ for $z\in \C$ around $z=0$.
Using the methods in \cite{Anisotropic,yang2019spiked}, we find the asymptotic limit, say $R_\infty(z)$, of $R(z)$ for any $z$ as $p\to \infty$ with an almost optimal convergence rate. In particular, when $z=0$, $\tr[\Sigma_2^{-1/2}A\Sigma_2^{-1/2}R_\infty(0)]$ gives the limit in \eqref{lem_cov_shift_eq}. 
On the other hand, we can write 
$$\bignorm{\Sigma_2^{1/2} (X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\beta}^2= \beta^\top \Sigma_2^{-1/2}R'(0)\Sigma_2^{-1/2} \beta.$$ 
Hence its limit can be calculated through $R_\infty'(z)$, which gives the expression in \eqref{lem_cov_derv_eq}. The details can be found in Appendix \ref{sec_maintools}. 
%The proof of Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative} is a main focus of Section \ref{sec_maintools}. 
We remark that one can probably derive the same asymptotic result using free probability theory (see e.g. \cite{nica2006lectures}), but our results \eqref{lem_cov_shift_eq} and \eqref{lem_cov_derv_eq} also give an almost sharp error bound $\bigo{ p^{-1/2+\epsilon}}$.

%We now state several helper lemmas to get estimates on $L(\hat{\beta}_t^{\STL})$ and $L(\hat{\beta}_t^{\MTL})$ for $t=2$. 
The next lemma, which is a folklore result in random matrix theory, helps to determine the asymptotic limit of $\te(\hat{\beta}_t^{\STL})=\sigma^2   \bigtr{(X_2^{\top}X_2)^{-1}\Sigma_2}$ as $p\to \infty$. When the entries of $X_2$ are multivariate Gaussian, this lemma recovers the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}. For general non-Gaussian random matrices, it can be obtained using Stieltjes transform method; see e.g., Lemma 3.11 of \cite{bai2009spectral}. Here we shall state a result obtained from Theorem 2.4 in \cite{isotropic}, which gives an almost sharp error bound.
\begin{lemma}\label{lem_minv}
Suppose $X$ satisfies assumption \ref{assm_secA1}. Let $A$ be any $p\times p$ matrix that is independent of $X$. We have that for any constant $\e>0$,
	\be\label{XXA}  \bigtr{(X_2^{\top}X_2)^{-1}A} = \frac{1}{\rho_2 - 1} \frac1p\tr(\Sigma^{-1}A) +\bigo{ \|A\|p^{-1/2+\epsilon}} \ee
with high probability.
\end{lemma}
Note that Lemma \ref{lem_cov_shift} extends Lemma \ref{lem_minv} for a single sample covariance matrix to the sum of two independent sample covariance matrices. Note that if we set $n_1=0$ and $n_2=n$, then $a_1 = 0$ and $a_2 = (n_2-p) / n_2$ is the solution to \eqref{eq_a12extra}, and one can see that \eqref{lem_cov_shift_eq} reduces to \eqref{XXA}.

>>>>

Based on the above intuition, in the following we develop generalization bounds that relate the bias and variance of multi-task learning to the sample sizes and the covariance matrices of both tasks.
We introduce a key quantity $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ that captures the covariate shift between task $1$ and task $2$.

@@

\begin{theorem}[Two-task case]\label{thm_main_informal}
	For the setting of two tasks, let $\delta > 0$ be a fixed error margin, $\rho_2 > 1$ and $\rho_1 \gtrsim \delta^{-2}\cdot \lambda_{\min}(M)^{-4} \norm{\Sigma_1} \max(\norm{\beta_1}^2, \norm{\beta_2}^2)$.
	Let $B, W_1, W_2$ be any local minimum of equation \eqref{eq_mtl}.
 	There exist two deterministic functions $\Delta_{\bias}$ and $\Delta_{\vari}$ that only depend on scaled model distance $\beta_1 - \frac{W_1}{W_2} \beta_2$, sample sizes $n_1 = \rho_1 \cdot p, n_2 = \rho_2 \cdot p$, and covariate shift matrix $M$ such that
	\begin{enumerate}
		\item[a)] \textbf{Positive transfer:} If $\Delta_{\bias} < \Delta_{\vari} -\delta$, then w.h.p. over the randomness of $X_1, X_2$, we have
			\[ \te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).  \]
		\item[b)] \textbf{Negative transfer:} If $\Delta_{\bias} > \Delta_{\vari} + \delta$, then w.h.p. over the randomness of $X_1, X_2$, we have
			\[ \te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}). \]
	\end{enumerate}
\end{theorem}

@@

\paragraph{Key ingredients.}
We shall refer to random matrices of the form $X^\top X$ as sample covariance matrices following the standard notations in high-dimensional statistics. The following two lemmas  are the main random matrix theoretical results of this paper, which will be used to estimate the two terms on the right-hand side of \eqref{eq_te_mtl_2task}.
Let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M$ in descending order.
%which deals with the inverse of the sum of two random matrices, which
%any is can be viewed as a special case of Theorem \ref{thm_model_shift}.

\begin{lemma}[Variance asymptotics]\label{lem_cov_shift}
	%Let $X_i\in\real^{n_i\times p}$ be a random matrix that contains i.i.d. row vectors with mean $0$ and variance $\Sigma_i\in\real^{p\times p}$, for $i = 1, 2$.
	%Suppose $X_1=Z_1\Sigma_1^{1/2}\in \R^{n_1\times p}$ and $X_2=Z_2\Sigma_2^{1/2}\in \R^{n_2\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho_1:=n_1/p>1$ and $\rho_2:=n_2/p>1$ being fixed constants.
	%Denote by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and
	In the setting of Theorem \ref{thm_main_informal}, let $\Sigma$ be any $p\times p$ fixed and deterministic matrix.
	We have that %for any constant $\e>0$,
	%When $n_1 = c_1 p$ and $n_2 = c_2 p$, we have that with high probability over the randomness of $X_1$ and $X_2$, the following equation holds
	\begin{align}\label{lem_cov_shift_eq}
		\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} \Sigma} = \frac{1}{n_1 + n_2}\cdot \bigtr{ (a_1 \Sigma_1 + a_2\Sigma_2)^{-1} A} + o(1)%\bigo{\|\| p^{-1/2+\epsilon}}
	\end{align}
with high probability, where $(a_1, a_2)$ is the solution to the following deterministic equations:
	\begin{align}
		a_1 + a_2 = 1- \frac{1}{\rho_1 + \rho_2},\quad a_1 + \frac1{\rho_1 + \rho_2}\cdot \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2} = \frac{\rho_1}{\rho_1 + \rho_2}. \label{eq_a12extra}
	\end{align}
\end{lemma}

Lemma \ref{lem_cov_shift} allows us to get a tight bound on equation \eqref{eq_te_var}, that only depends on \textit{sample size}, \textit{covariate shift} and the scalar $\hat{v}$.
As a remark, the concentration error $\bigo{\|A\| p^{-1/2+\epsilon}}$ of our result is nearly optimal.
The next lemma, which is a folklore result in random matrix theory, helps to determine the asymptotic limit of $\te(\hat{\beta}_t^{\STL})=\sigma^2   \bigtr{(X_2^{\top}X_2)^{-1}\Sigma_2}$ as $p\to \infty$. When the entries of $X_2$ are multivariate Gaussian, this lemma recovers the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}. For general non-Gaussian random matrices, it can be obtained using Stieltjes transform method; see e.g., Lemma 3.11 of \cite{bai2009spectral}. Here we shall state a result obtained from Theorem 2.4 in \cite{isotropic}, which gives an almost sharp error bound.
\begin{corollary}[e.g. Theorem 2.4 in \citet{isotropic}]\label{lem_minv}
	In the setting of Lemma \ref{lem_cov_shift}, with high probability we have that
	\be\label{XXA}  \bigtr{(X_2^{\top}X_2)^{-1}A} = \frac{1}{n_2 - p} \cdot \bigtr{\Sigma_2^{-1}A} + o(1). \ee
\end{corollary} %\bigo{ \|A\|p^{-1/2+\epsilon}}
Note that Lemma \ref{lem_cov_shift} extends Lemma \ref{lem_minv} for a single sample covariance matrix to the sum of two independent sample covariance matrices. Note that if we set $n_1=0$ and $n_2=n$, then $a_1 = 0$ and $a_2 = (n_2-p) / n_2$ is the solution to \eqref{eq_a12extra}, and one can see that \eqref{lem_cov_shift_eq} reduces to \eqref{XXA}.

For the bias term of equation \eqref{eq_te_model_shift}, a similar result that scales with task model distance in addition to sample size and covariate shift holds (cf. Lemma \ref{lem_cov_derivative} in Appendix \ref{app_proof_main_thm}).

\begin{lemma}[Bias asymptotics]\label{lem_cov_derivative}
In the setting of Theorem \ref{thm_main_informal}, let $\beta \in \R^p$ be any vector that is independent of $X_1$ and $X_2$. We have that for any constant $\e>0$,
\begin{equation}\label{lem_cov_derv_eq}
\begin{split}
&(n_1+n_2)^2\bignorm{\Sigma_2^{1/2} (\hat v^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\beta}^2 \\
&= \beta^{\top} \Sigma_2^{-1/2}  \frac{(1 + a_3)\id + \hat v^2a_4 {M}^{\top}{M}}{( \hat v^2a_1 {M}^{\top}{M}+a_2)^2} \Sigma_2^{-1/2} \beta +\OO(p^{-1/2+\e}\|\beta\|^2),
\end{split}
\end{equation}
with high probability, where $a_{3}$ and $a_4$ satisfy the following system of linear equations:
\begin{gather}\label{eq_a34extra}
		\left(\rho_2  a_2^{-2}-  b_0\right)\cdot  a_3 - b_1 \cdot  a_4
		= b_0, \quad \left(\rho_1 a_1^{-2} -  b_2  \right)\cdot  a_4 -  b_1 \cdot  a_3 = b_1 .
%		\left(\frac{n_1}{\hat a_1^2} -  \sum_{i=1}^p \frac{\hat \lambda_i^4   }{  (\hat a_2 + \hat \lambda_i^2\hat a_1)^2  }\right)\hat a_4 -\left(\sum_{i=1}^p \frac{\hat \lambda_i^2  }{  (\hat a_2 + \hat \lambda_i^2\hat a_1)^2  }\right)\hat a_3
%		= \sum_{i=1}^p \frac{\hat \lambda_i^2 }{  (\hat a_2 + \hat \lambda_i^2\hat a_1)^2  }. \label{eq_a4}
	\end{gather}
Here $b_0$, $b_1$ and $b_2$ are defined as
$$ b_k:= \frac1{p}\sum_{i=1}^p \frac{ \hat v^{2k}\lambda_i^{2k}}{ ( a_2 + \hat v^2\lambda_i^2 a_1)^2  },\quad k=0,1,2.$$
\end{lemma}

\medskip
\noindent\textbf{Proof overview.}
%We first describe the proof of Theorem \ref{lem_cov_shift_informal}.
The proofs of Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative} are based on the Stieltjes transform method (or the resolvent method) in random matrix theory \cite{bai2009spectral,tao2012topics,erdos2017dynamical}. Roughly speaking, we study the resolvent $R(z):=[\Sigma_2^{-1/2}( X_1^{\top}X_1 + X_2^{\top}X_2)\Sigma_2^{-1/2}-z]^{-1}$ for $z\in \C$ around $z=0$.
Using the methods in \cite{Anisotropic,yang2019spiked}, we find the asymptotic limit, say $R_\infty(z)$, of $R(z)$ for any $z$ as $p\to \infty$ with an almost optimal convergence rate. In particular, when $z=0$, $\tr[\Sigma_2^{-1/2}A\Sigma_2^{-1/2}R_\infty(0)]$ gives the limit in \eqref{lem_cov_shift_eq}. 
On the other hand, we can write 
$$\bignorm{\Sigma_2^{1/2} (X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\beta}^2= \beta^\top \Sigma_2^{-1/2}R'(0)\Sigma_2^{-1/2} \beta.$$ 
Hence its limit can be calculated through $R_\infty'(z)$, which gives the expression in \eqref{lem_cov_derv_eq}. The details can be found in Appendix \ref{sec_maintools}. 
%The proof of Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative} is a main focus of Section \ref{sec_maintools}. 
We remark that one can probably derive the same asymptotic result using free probability theory (see e.g. \cite{nica2006lectures}), but our results \eqref{lem_cov_shift_eq} and \eqref{lem_cov_derv_eq} also give an almost sharp error bound $\bigo{ p^{-1/2+\epsilon}}$.

%We now state several helper lemmas to get estimates on $L(\hat{\beta}_t^{\STL})$ and $L(\hat{\beta}_t^{\MTL})$ for $t=2$.
